Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-10-22 paper/main.tex
add notes for christian to read
restructure mechanics and interactions section

The mechanical behaviour of rod-shaped bacteria is governed by a plethora of effects.
In this work, we aim to establish a model which is general enough such that it can be reused as the
basis for many variants of rod-shaped bacteria but also explicit enough such that we can quantify
its parameters.
We thus utilize only a minimal set of assumptions.

When describing the bacterial system, we distinguish between intra-cellular processes and cell-cell
interactions.
We restrict ourselves to the features discussed below and summarize them in
Table~\ref{table:simulation-aspects}.
% For a collection of other known aspects which could be considered, we refer to the supplementary
% section~\ref{section:discussion}, and~\ref{section:supplement-more-simulation-aspects}.

@@

% --------------------------------------------------------------------------------------------------
\subsection{Spatial Representation}
\label{subsec:spatial-representation}
This work makes use of a few but powerful assumptions concerning the mathematical representation of
the cells. In order to describe the spatial dynamics of individual cells or agents, we need to
define their position, shape and how they interact with each other.

In order to provide as much flexibility as possible when constructing new models, we restrict
ourselves to assumptions concerning the position such that researchers can freely experiment with
various shapes and interaction types.
\textbf{[Dieser Satz ist leider unverständlich. Je öfter ich den Satz lese, umso weniger ist mir klar, was du sagen möchtest.]}\\

Furthermore, it is very important to strike a good balance between the restrictiveness of our
assumptions and generality in order to provide meaningful results and still be able to apply our
approach to a broader class of models.
\textbf{[Ebenso unklar warum du das hier an dieser Stelle sagst. Vielleicht versuchst du die Aussagen zu bündeln und geeigneter Stelle zu bringen. So ist man leider eher verwirrt.]}\\

Taking into account the natural shape of rod-like bacteria, it is clear that they can not be
represented with a singular point $\vec{x}$ in space.
\textbf{[Hast du nicht bereits oben die Hilfsvariablen $x_i$ eingeführt?]}\\

We could extend this positional definition with a vector $\vec{w}$, indicating the direction of
extension but this would restrict ourselves to fully straight rod-shaped bacteria.
\textbf{[Nicht spekulieren was du hättest machen können, sondern einfach sagen, was du gemacht hast.]}

However, multiple experiments have shown that MReB is responsible for the growth and maintenance of
the elongated shape of bacteria~\cite{Erickson2001}.
\textbf{[Wiederholung?]}

By means of this protein, rod-shaped bacteria can bend elastically on short time-scales and deform
plastically over longer periods of time. Our goal is to provide a model that can encapsulate these effects.
\textbf{[Das beschriebene Verhalten ist visko-elastisch. Soweit ich sehe, hast du das nicht implementiert. Könnte es sein, dass das elastisches Modell mit Wachstum äquivalent zum Maxwell-Modell ist? Eigene Antwort: im allgemeinen nein.]}\\

We discretize the (possibly non-straight) rod of a bacteria and model it by using multiple vertices
which are part of a 1-dimensional
polygon with two endpoints $\vec{x}_i$. \textbf{[Was ist denn die Diskretisierungslänge? Two endpoints $x_i$ macht keinen Sinn.]}\\

Figure~\ref{fig:mechanics-bacterium} shows a visualization of this approach.
The angles $\alpha_i$ between vertices $\vec{x}_{i-1},\vec{x}_i$ can be used when modeling the
dynamics of these bacteria.
\textbf{[Klingt kryptisch. "Can be used ..". Für was, wie?]}\\

With these fundamental thoughts layed out, we retain the flexibility to represent many models such
as varying shapes since we do not have to make any assumptions on the type of interactions between
two bacteria.
Even concerning the shape of bacteria, this assumption still allows us to model elliptical cells by
modifying the interaction type as well as other effects such as polarity of
interactions~\cite{Duvernoy2018}.
Furthermore, we also did not have to assume any particular mechanical dynamics of the agents, only
their spatial representation, thus leaving in particular room for plastical or elastic deformations
of the rod and other more intricate mechanisms.
\textbf{[Ich glaube, ich verstehe warum du das hier sagst. Aber das kommt nicht richtig rüber. Der Absatz ist vielleicht nicht nötig. Der Kommentar über elliptische Zellen, sollte eher unter Potential stehen?]}\\

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{docs/source/_static/mechanics.png}
    \caption{
        Example of an elonged bacterial rod represented by a collection of vertices which are
        connected by springs.
        The angles $\alpha_i$ have been exaggerated such that they can be identified more easily.
        A bacterial agent which is not interacting with any external objects will tend to the
        equilibrium state of a straight rod with no curvature.
    }
    \label{fig:mechanics-bacterium}
\end{figure}

% --------------------------------------------------------------------------------------------------
\subsection{Mechanics}
\label{subsection:mechanical-model-mechanics}
We use principles from discrete differential geometry~\cite{bobenko2008discrete} to describe the
stiffness of the bacterial rod~\cite{Amir2014_2}.
As explained above, MReB will counteract the bending of the rod, resulting in a stiffening force.
In principle, this stress will be dependent on the position inside the cell~\cite{Chatterjee1988},
but for sake of simplicity we chose to model the bacterium as a bending beam.
\textbf{[", thus utilizing the simplest solution which can still produce the desired effects." Es ist unklar was die "desired effects" sind. ]}\\

Such a beam can be represented by a collection of vertices $\vec{x}_i$ which are connected by a
springs of lengths $l_i$.
\textbf{[Das klingt nach einer Wiederholung von Spatial Representation. Du erklärst ziemlich umständlich. Konzentriere dich auf das wesentliche und überlege, was du eigentlich sagen möchtest.]}\\

We use the notation $\vec{c}_i:=\vec{x}_i-\vec{x}_{i-1}$ with which the resulting force acting
between two vertices if given by

\begin{align}
    \vec{F}_{i,\text{springs}} =
        &-\gamma\left(1 - \frac{l}{\left|\vec{c}_i\right|}\right)
        \vec{c}_i\\
        &+ \gamma\left(1 - \frac{l}{\left|\vec{c}_{i+1}\right|}\right)
        \vec{c}_{i+1}.
\end{align}

The value of the spring constants determines the elasticity with respect to elongation of the cell.
In situations where this aspect is not relevant, they merely serve as a auxiliary parameters to
maintain the length of the rod.
In addition to springs between individual vertices $\vec{x}_i$, we assume that each angle at a
vertex between edges is subject to a force indiced by curvature.
% We assume that we can model the mechanical properties of the bacterium as an elastic rod.
Given the angle $\alpha_i = \sphericalangle(\vec{c}_{i-1},\vec{c}_i)$ between adjacent edges,
the bending force is proportional to the curvature $\kappa_i$ at each vertex
$\kappa_i = 2\tan(\alpha_i/2)$.
The resulting force acts along the angle bisector which can be calculated from the edge vectors.
Figure~\ref{fig:mechanics-bacterium} shows an elongated rod represented by $6$ vertices.
The forces acting on vertices $\vec{x}_i,\vec{x}_{i-1},\vec{x}_{i+1}$ are given by

\begin{align}
    \vec{F}_{i,\text{curvature}} &= \eta\kappa_i
        \frac{\vec{c}_i - \vec{c}_{i+1}}{|\vec{c}_i-\vec{c}_{i+1}|}\\
    \vec{F}_{i-1,\text{curvature}} &= -\frac{1}{2}\vec{F}_{i,\text{curvature}}\\
    \vec{F}_{i+1,\text{curvature}} &= -\frac{1}{2}\vec{F}_{i,\text{curvature}}
\end{align}

where $\eta_i$ is the rigidity at vertex $\vec{x}_i$ (see Figure~\ref{fig:mechanics-bacterium}).
We can see that the curvature force does not move the overall center of the rod in space.
The total force is the sum of external and interal forces.

@@

and are integrated via damped brownian dynamics

\begin{align}
    \partial_t^2 \vec{x}_i &= \partial_t\vec{x}_i + \sqrt{2D}\vec{\xi}_i\\
    \partial_t\vec{x}_i &= \vec{F}_{i,\text{total}} - \lambda \partial_t\vec{x}_i
\end{align}
\textbf{[Ich verstehe die Gleichungen nicht. Eine überdämpfte Langevin Gleichung bezeichnet man doch als Brownsche Bewegung, oder? Und sollte die Brownsche Gleichung nicht so lauten?
\begin{align}
   \partial_t \vec{x}_i &= \frac{D}{k_B T} \vec{F}_{i,\text{total}} + \sqrt{2D}\vec{\xi}_i
\end{align}
Die Einheiten stimmen auch überhaupt nicht. $[x_i]=$m, $[\lambda]=$kg/s, $[F]=$(kg m)/s$^2$
]}\\

where $D$ is the diffusion constant, $\vec{\xi}$ is a Wiener process and $\lambda$ a damping constant.
Together they determine the stochastic motion of the individual vertices which can arise from
recurring collisions with small particles of the surrounding medium or other minor fluctuations.

@@

Calculating the force between cells resulting  from long ranged interactions requires performing a
double integral over the length of the cells. 
To reduce the computational costs we use a approximative, simplified model Given a vertex
$\vec{x}_i$ on one cell, we determine the closest point $\vec{p}$ on the polygonal line given by the
vertices $\{\vec{w}_j\}$ of the interacting cell. 
\textbf{[Ich bin etwas verwirrt. Warum änderst du die Bezeichnung der Vertices?]}

Furthermore we determine the value $q\in[0,1]$ such that $\vec{p} = (1-q)\vec{w}_j + q\vec{w}_{j+1}$
for some specific $j$.
The force is then calculated between the points $\vec{x}_i$ and $\vec{p}_i$ and acts on
the vertex $\vec{w}_i,\vec{w}_{i+1}$ with relative strength $(1-q)$ and $q$.

@@

The interaction between cells can be described by the Mie potential~\cite{Mie1903}.
It is the generalization of the well-known Lennard-Jones~\cite{Jones1924} potential and was designed
to describe the interactions between molecules.
The Mie potential is given by:

\begin{align}
    U(r) &= C\epsilon\left[ \left(\frac{\sigma}{r}\right)^n -
        \left(\frac{\sigma}{r}\right)^m\right]\\
    C &= \frac{n}{n-m}\left(\frac{n}{m}\right)^{\frac{n}{n-m}}\\
    V(r) &= \min(U(r), \beta)\theta(r-\zeta)
\end{align}

The strength of the potential is given by $\epsilon$ and the radius \textbf{[Welcher Radius?]} is
encoded in $\sigma = (m/n)^{1/(n-m)}(R_1+R_2)$ where $R_1,R_2$ are the radii of the two interacting
particles.
In our simulation, we use $V(r)$ to guarantee numerical regularity.
\textbf{[Beschreibe, welche Rolle $\beta$ und $\zeta$ spielen.]}

The specific choice of interaction potential plays a crucial role for the dynamics of the agents.
Since we cannot expect to simply know the correct shape and parameters of this potential, it is
desirable to experiment with various interaction potentials in order to quantify their
ability to produce accurate results.
While this study, only relies on the Mie Potential, our python package \texttt{cr\_mech\_coli}
also implements the popular Morse potential~\cite{Morse1929} to model physical interactions.

\textbf{[Du hast hier von shape geredet. Das Potential ist doch immer radial-symmetrisch, oder? Vielleicht könnte es helfen, Bilder zu zeigen? Darin könnte man die Bedeutung der Parameter verdeutlichen. Wenn du schreibst, das Potential ist sehr wichtig, dann musst du viel besser begründen, warum du das Potential gewählt hast. Ich denke, du solltest das etwas abschwächen.]}

@@

Afterwards, we start at one endpoint $\vec{x}_0$ and move along the polygon defined by the
bacterial rod. We stop when we have reached a length that matches the length of the new spring length $\tilde{l}$.
From here, we continue and iterate this procedure until we have found all new vertices $\vec{w}_i$.
Equation~\ref{eq:growth-rod-new-lengths} ensures that that the tips of the two new rods are touching
right after the division event.
\textbf{[Was ist $r$? Was ist mit den Parametern und mit anderen intro-zellularen Zuständen?]}

>>>>

@Christian
\begin{itemize}
    \item shortened everything
    \item leave out parts which are not directly used by the model
    \item move biological context to respective subsections to link more directly to model
    \item remove duplicates and useless text
    \item move heterogeneity considerations from table in main text
\end{itemize}

The mechanical behaviour of rod-shaped bacteria is governed by a plethora of effects.
In this work, we aim to establish a model which is general enough such that it can be reused as the
basis for many variants of rod-shaped bacteria but also explicit enough such that we can quantify
its parameters.
We thus utilize only a minimal set of assumptions.
When describing the bacterial system, we distinguish between intra-cellular processes and cell-cell
interactions.
We restrict ourselves to the features discussed below and summarize them in
Table~\ref{table:simulation-aspects}.
% For a collection of other known aspects which could be considered, we refer to the supplementary
% section~\ref{section:discussion}, and~\ref{section:supplement-more-simulation-aspects}.

@@

% --------------------------------------------------------------------------------------------------
\subsection{Spatial Representation}
\label{subsec:spatial-representation}
This work makes use of a one powerful assumption concerning the mathematical representation of the
cells.
To model the spatial mechanics of elongated bacteria \cite{Billaudeau2017}, we represent them as a
collection of auxiliary vertices $\{\vec{x}_i\}$.
We treat this collection of vertices as a discretization of the rod where the individual segments
between the vertices carry no explicit physical or biological meaning.
This assumption alone is enough to be able to build visualization and data-extraction pipelines (see
\ref{fig:flowchart-project-structure}).
Our computational model builds on top of this assumption and utilizes this representation to
define the dynamics of the vertices $\{\vec{x}_i\}$.
For this study, we will choose a particular type of model incorporating a chosen set effects.
But researchers are in general free to adapt the model for the specific problem at hand.
By building on our discretization scheme, future modeling approaches will be able to directly reuse
our established theoretical methods and computational functionalities.

\textbf{TODO mention this in sections below}
\begin{itemize}
    \item plastic/elastic deformations
    \item non-uniform thickness of the rod
    \item polar interactions \cite{Duvernoy2018}
    \item other dynamics
\end{itemize}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{docs/source/_static/mechanics.png}
    \caption{
        Example of an elonged bacterial rod represented by a collection of vertices which are
        connected by springs.
        The angles $\alpha_i$ will be used to model the rigidity of the rod.
        A bacterial agent which is not interacting with any external objects will tend to the
        equilibrium state of a straight rod with no curvature.
    }
    \label{fig:mechanics-bacterium}
\end{figure}

% --------------------------------------------------------------------------------------------------
\subsection{Mechanics}
\label{subsection:mechanical-model-mechanics}
In rod-shaped bacteria, the MreB protein plays a key role for the maintenance of the cellular shape.
It forms an actin-like cytoskeleton~\cite{Erickson2001,Dersch2020} and localizes specifically at
regions of negative curvature thus introducing heterogeneous cell growth~\cite{Ursell2014} which
results in a bending force during growth.
This feedback mechanism will result in straight rod-shaped bacteria if not interfered from the
external environment~\cite{Wang2010}.

We use principles from discrete differential geometry~\cite{bobenko2008discrete} to describe the
stiffness of the bacterial rod~\cite{Amir2014_2}.
As explained above, MReB will counteract the bending of the rod, resulting in a stiffening force.
We model the bacterium as a bending beam.
% In principle, this stress will be dependent on the position inside the cell~\cite{Chatterjee1988},
% but for sake of simplicity we chose to model the bacterium as a bending beam.
% \textbf{[", thus utilizing the simplest solution which can still produce the desired effects." Es ist unklar was die "desired effects" sind. ]}\\
Using our discretization $\vec{x}_i$, we connect the vertices by springs of length $l_i$.
% which can be represented by a collection of vertices $\vec{x}_i$ which are connected by a
% springs of lengths $l_i$.
% \textbf{[Das klingt nach einer Wiederholung von Spatial Representation. Du erklärst ziemlich umständlich. Konzentriere dich auf das wesentliche und überlege, was du eigentlich sagen möchtest.]}\\
We use the notation $\vec{w}i:=\vec{x}_i-\vec{x}_{i-1}$ with which the resulting force acting
between two vertices if given by

\begin{equation}
    \vec{F}_{i,\text{springs}} =
        -\gamma\left(1 - \frac{l}{\left|\vec{w}i\right|}\right) \vec{w}i
        + \gamma\left(1 - \frac{l}{\left|\vec{w}{i+1}\right|}\right) \vec{w}{i+1}.
\end{equation}

The value of the spring constants determines the elasticity with respect to elongation of the cell.
In situations where this aspect is not relevant, they merely serve as a auxiliary parameters to
maintain the length of the rod.
In addition to springs between individual vertices $\vec{x}_i$, we assume that each angle at a
vertex between edges is subject to a force indiced by curvature.
% We assume that we can model the mechanical properties of the bacterium as an elastic rod.
Given the angle $\alpha_i = \sphericalangle(\vec{w}{i-1},\vec{w}i)$ between adjacent edges,
the bending force is proportional to the curvature $\kappa_i$ at each vertex
$\kappa_i = 2\tan(\alpha_i/2)$.
The resulting force acts along the angle bisector which can be calculated from the edge vectors.
Figure~\ref{fig:mechanics-bacterium} shows an elongated rod represented by $6$ vertices.
The force acting on vertex $\vec{x}_i$ is given by

\begin{equation}
    \vec{F}_{i,\text{curvature}} = \eta\kappa_i
        \frac{\vec{w}i - \vec{w}{i+1}}{|\vec{w}i-\vec{w}{i+1}|}
\end{equation}

where $\eta_i$ is the rigidity at vertex $\vec{x}_i$ (see Figure~\ref{fig:mechanics-bacterium}).
The adjoining vertices $\vec{x}_{i-1},\vec{x}_{i+1}$ are subject to half of the same force in the
opposite direction $\vec{F}_{i\pm 1,\text{curvature}} = -0.5\vec{F}_{i,\text{curvature}}$.
We can see that the curvature force does not move the overall center of the rod in space.
The total force is the sum of interal and external forces.

@@

and are integrated via damped newtonian dynamics

\begin{equation}
    \partial_t^2 \vec{x}_i = \vec{F}_{i,\text{total}} - \lambda \partial_t\vec{x}_i
\end{equation}

where $\lambda$ a damping constant and the mass was set to $1$ to simplify the model.

@@

Calculating the force between cells requires performing a double integral over the length of the
cells.
To reduce the computational costs we use an approximative, simplified model.
Given a vertex $\vec{x}_i$ on one cell, we determine the closest point $\vec{p}$ on the polygonal
line given by the vertices $\{\vec{y}_j\}$ of the interacting cell.
Furthermore we determine the value $q\in[0,1]$ such that $\vec{p} = (1-q)\vec{y}_j + q\vec{y}_{j+1}$
for some specific $j$.
The force is then calculated between the points $\vec{x}_i$ and $\vec{p}$ and acts on
the vertex $\vec{y}_j,\vec{y}_{j+1}$ with relative strength $(1-q)$ and $q$.

@@

We utilize the Mie potential~\cite{Mie1903} to describe the attraction between two cells.
It is the generalization of the well-known Lennard-Jones~\cite{Jones1924} potential and was designed
to describe the interactions between molecules.
The Mie potential is given by:

\begin{align}
    U(r) &= C\epsilon\left[ \left(\frac{\sigma}{r}\right)^n -
        \left(\frac{\sigma}{r}\right)^m\right]\\
    C &= \frac{n}{n-m}\left(\frac{n}{m}\right)^{\frac{n}{n-m}}
\end{align}

The strength of the potential is given by $\epsilon$ and the thickness of our rods
is encoded in $\sigma = (m/n)^{1/(n-m)}(R_1+R_2)$ where $R_1,R_2$ are the thicknesses of the two
interacting bacteria.
We utilize the shorthand notation $r=|\vec{x}_i - \vec{p}|$ for the distance between the two points
in question.
We numerically regularize the potential by imposing an upper bound $\beta$ on the norm of the
resulting force $\vec{F} = \vec{\nabla} V$.
Furthermore, we impose a cutoff parameter $\zeta$ which sets sets the force to zero when the
distance $r$ between the interacting points exceeds this threshold.
The resulting force is given by

\begin{equation}
    \vec{F}_{i,\text{external}} = \vec{F} \theta(r-\zeta) \min\left(1, \beta / |\vec{F}|\right)
\end{equation}

where $\theta$ is the heaviside step-function and the $\min$ term ensures the upper bound $\beta$.

The physical interaction between the bacteria plays a crucial role for the overall dynamics of our
simulation.
The Mie potential is capable of describing a wide range of interaction types from soft to rigid, by
adjusting the parameters of the potential accordingly.
In contrast, a popular alternative is the Morse potential~\cite{Morse1929}, which is less
customizable due to the lower amount of parameters and more restrictive mathematical formulation.
Therefore, the Mie potential is a good choice for a simple, interpretable interaction potential with
low amount of parameters that can be estimated by a data-driven approach.

@@

Afterwards, we start at one endpoint $\vec{x}_0$ and move along the polygon defined by the
bacterial rod. We stop when we have reached a length that matches the length of the new spring length $\tilde{l}$.
From here, we continue and iterate this procedure until we have found all new vertices $\vec{y}_i$.
Equation~\ref{eq:growth-rod-new-lengths} ensures that that the tips of the two new rods are touching
right after the division event.
\textbf{[Was ist $r$? Was ist mit den Parametern und mit anderen intro-zellularen Zuständen?]}
